Palaeodistribution of crop progenitors and wild foods in the Southern Levant

### 2023-08-16 analysis/analysis.Rmd
Refactor occurrence reading and tidying

```{r definitions}
latlong <- 4326
utm37n <- 32637

@@

@@

```{r data-basemap}
# Turn off s2 while we deal with Natural Earth's dodgy geometries
use_s2 <- sf_use_s2(FALSE)

@@

s_levant_land <- st_intersection(st_as_sfc(s_levant), ne_land)

@@

Occurrence data was obtained from GBIF.

```{r data-occ}
gbif_data <- function(scientificName, ...) {
  gbif_data <- rgbif::occ_data(scientificName = scientificName, ...)
  gbif_data$data
}

gbif_path <- here::here("analysis", "data", "derived_data", "gbif")

taxa <- c("Triticum dicoccoides" = "Triticum dicoccoides",
          "Triticum urartu" = "Triticum urartu",
          "Hordeum spontaneum" = "Hordeum spontaneum",
          "Bolboshoenus spp." = "Bolboschoenus")

# Get occurrence data from GBIF
# Including only records with coordinates and no recorded issues with geospatial
#   data.
taxa |>
  map(gbif_data, hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
      geometry = w_asia, limit = 100000) |>
  list_rbind(names_to = "taxon") ->
  occ

# Exclude:
# * Coordinate uncertainty of 5 km or more (≈2.5 mins)
# * Fossil occurrences
occ |>
  filter(coordinateUncertaintyInMeters < 5000 | is.na(coordinateUncertaintyInMeters)) |>
  filter(basisOfRecord != c("FOSSIL_SPECIMEN")) ->
  occ

# Remove redundant observations of the same taxon in the same place
occ |>
  group_by(taxon) |>
  distinct(decimalLongitude, decimalLatitude, .keep_all = TRUE) |>
  ungroup() ->
  occ

# Drop unneeded columns, normalise column names and convert to sf
occ |>
  select(
    taxon, gbif_key = key, genus, species,
    longitude = decimalLongitude, latitude = decimalLatitude
  ) |>
  st_as_sf(coords = c("longitude", "latitude"), 
           crs = 4326, remove = FALSE) ->
  occ
```

```{r plot-occ-distribution}
ggplot(occ) + 
  facet_wrap(vars(taxon)) +
  layer_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(s_levant, fill = alpha("red", 0.15), colour = "red") +
  layer_spatial(occ, alpha = 0.2) +
  layer_spatial(ne_countries, fill = NA) +
  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
           ylim = w_asia[c("ymin", "ymax")],
           crs = utm37n,
           default_crs = latlong)
```

@@

```{r data-background}
n_background <- 10000

occ |>
  mutate(present = TRUE) |>
  group_by(taxon) |>
  nest(.key = "occ_data") |>
  mutate(bg_data = map(1:n(), \(i) { 
    background_sample(
      region = s_levant_land, 
      N = n_background,
      coord_x = "longitude", coord_y = "latitude",
      present = FALSE
    )
  })) |>
  mutate(data = map2(occ_data, bg_data, bind_rows)) |>
  select(taxon, data) |>
  unnest(data) |>
  st_sf() ->
  occ
```

### Climate data

@@

### Other predictor data

Soil, terrain (incl. TWI, slope?)

@@

```{r model}
rf_pred_vars <- function() {
  c(any_of("present"), starts_with("bio"))
}

# Drop any rows with NA values in any of the predictor variables
# TODO: Where do these actually come from? There's usually only a handful.
occ <- drop_na(occ, rf_pred_vars())

@@

rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger",
    probability = TRUE,
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1)
  ) -> 
  rf_model

# Fit to each taxon
occ |>
  nest(.by = taxon) |>
  mutate(
    data = map(data, as_tibble),
    fit = map(data, \(data) {
      fit(rf_model, factor(present) ~ ., data = select(data, rf_pred_vars()))
    })) ->
  rf_fits
```

@@

```{r predict}
rf_fit <- rf_fits$fit[[1]]

# TODO: predict on all periods (maybe combine stars and collapse into dimension first?)
rf_pred_cur <- predict(climate$cur, rf_fit, type = "prob")
rf_pred_eh <- predict(climate$EH, rf_fit, type = "prob")
rf_pred_yds <- predict(climate$YDS, rf_fit, type = "prob")
rf_pred_ba <- predict(climate$BA, rf_fit, type = "prob")
rf_pred_hs1 <- predict(climate$HS1, rf_fit, type = "prob")
rf_pred_lgm <- predict(climate$LGM, rf_fit, type = "prob")

rf_pred <- c(rf_pred_cur, rf_pred_eh, rf_pred_yds, rf_pred_ba, rf_pred_hs1,
             rf_pred_lgm)
names(rf_pred) <- c("cur", "EH", "YDS", "BA", "HS1", "LGM")

ggplot() + 
  geom_stars(data = st_redimension(st_crop(rf_pred, w_asia))) + 
  facet_wrap(~new_dim)
```

### Verification

>>>>

```{r setup-parallel}
# Several parts of this analysis are processing- and memory-intensive.
# You may need to adjust the futures strategy specified here to fit your 
# computational resources. See ?future::plan for options.
plan("multisession")
```

```{r define-regions}
latlong <- 4326
utm37n <- 32637

@@

```{r define-periods}
archaeo_periods <- tribble(
  ~period,             ~agriculture,                          ~start_bp, ~end_bp,
  "Late Epipal.",      "Foraging",                            15000,     11700,
  "PPNA",              "Pre-domestication cultivation",       11700,     10700,
  "EPPNB",             "Cultivation of domesticated species", 10700,     10200,
  "MPPNB",             "Cultivation of domesticated species", 10200,     9500,
  "LPPNB/C",           "Agriculture",                         9500,      8500,
  "Pottery Neolithic", "Agriculture",                         8500,      6500,
  "Chalcolithic",      "Agriculture",                         6500,      5000,
)

neolithic <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C", "Pottery Neolithic")
```

@@

```{r read-basemap}
# Turn off s2 while we deal with Natural Earth's dodgy geometries
use_s2 <- sf_use_s2(FALSE)

@@

s_levant_land <- st_intersection(st_as_sfc(s_levant), ne_land)
w_asia_land <- st_intersection(st_as_sfc(w_asia), ne_land)

@@

```{r read-archaeo_flora}
archae_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv"))
```

```{r tidy-archaeo_flora}
archaeo_flora |>
  rename(taxon_group = "taxon", taxon = "taxon_detail") |>
  # Assign to archaeological periods
  # TODO: refactor as function
  # TODO: assign climate periods too
  mutate(
    age_mid = age_end + ((age_start - age_end) / 2),
    period = cut(-age_mid, 
                 breaks = -c(archaeo_periods$start_bp, 0), 
                 labels = archaeo_periods$period,
                 ordered_result = TRUE)
  ) ->
  archaeo_flora
```

```{r summarise-archaeo_flora}
n_assemb_neolithic <- archaeo_flora |> 
  distinct(site_name, phase_code, period) |> 
  filter(period %in% neolithic) |> 
  nrow()

n_site_neolithic <- archaeo_flora |> 
  filter(period %in% neolithic) |> 
  distinct(site_name) |> 
  nrow()
```

To generate the list of taxa modelled in this study, we ...

* Taxa present at more than three Neolithic sites, according to the database of our previous study [@ArranzOtaeguiRoe2023]
* Identified at least to species (and matchable to the GBIF Backbone Taxonomy)
* Have at least 100 occurrences in West Asia (meeting the criteria below) in the GBIF database
* Excluded domestic species, replacing them with their wild progenitor, where known and present in the GBIF database

We also considered the number modern occurrences available for each taxon.
ENMs are known to have reasonable predictive power even with small training samples (i.e. less than a hundred observations, [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008]).

```{r generate-flora}
archaeo_flora |>
  filter(period %in% neolithic) |>
  group_by(taxon) |>
  summarise(
    n_present = length(unique(phase_code)),
    p_present = n_present / n_assemb_neolithic,
    .groups = "drop"
  ) |>
  filter(n_present > 2) ->
  flora
```

```{r tidy-flora-taxonomy}
# Normalise species names and match to the GBIF Backbone Taxonomy
# Excluding taxa not identified to species
flora |>
  filter(!str_ends(taxon, "sp.")) |>
  mutate(
    taxon = case_match(taxon,
                       "Vicia narbonense" ~ "Vicia narbonensis", # N=3
                       "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
                       "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
                       "Lens orientalis" ~ "Vicia orientalis", # N=2
                       "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
                       "Stipa capensis" ~ "Stipa dregeana", # N=2
                       .default = taxon),
    taxon_original = taxon,
    taxon = map_chr(taxon, gbif_accepted_name)
  ) ->
  flora
```

```{r tidy-flora-domesticates}
# Recode domesticates as their wild progenitors.
# Add other wild progenitors of wheats
flora |> 
  mutate(taxon = case_match(
    taxon,
    "Cicer arietinum" ~ "Cicer reticulatum",
    "Hordeum vulgare" ~ "Hordeum spontaneum",
    "Vicia lens" ~ "Vicia orientalis",
    "Linum usitatissimum" ~ "Linum bienne",
    "Triticum aestivum" ~ "Triticum turgidum dicoccum",
    "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
    "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
    "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
    .default = taxon
  )) |>
  bind_rows(tibble(taxon = c("Triticum urartu", "Aegilops speltoides", 
                             "Aegilops tauschii"))) ->
  flora
```

```{r fetch-occ-count}
flora |>
  group_by(taxon) |>
  summarise(
    taxon_original = list(c(taxon_original)),
    n_present = max(n_present), # TODO: be more precise?
    p_present = max(p_present)
    ) |>
  mutate(n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                               hasCoordinate = TRUE, 
                                               hasGeospatialIssue = FALSE,
                                               geometry = bbox_wkt(w_asia)))) ->
  flora
```

```{r tbl-occ-count}
#| tbl-cap: "Species modelled in this study."
# TODO: add to caption:
# - with presence in Neolithic sites in West Asia, out of the X sites included in Y
# - with number of reliably-georeferenced occurrences in West Asia recorded in GBIF
# TODO: add count of species included?
flora |>
  mutate(
    taxon_detail = map_chr(taxon_original, paste, collapse = ", "),
    taxon_detail = if_else(taxon == taxon_detail, NA_character_, taxon_detail),
    include = n_occ > 50
  ) |>
  select(-taxon_original) |>
  group_by(include) |>
  arrange(-n_present) |>
  gt() |>
  cols_merge(
    c(taxon, taxon_detail),
    rows = !is.na(taxon_detail),
    pattern = "{1}<br><small>(incl. {2})</small>"
  ) |>
  fmt_percent(p_present, decimals = 0) |>
  cols_merge_n_pct(n_present, p_present) |>
  tab_style(cell_text(style = "italic"), cells_body(taxon))
```

```{r filter-flora}
flora <- filter(flora, n_occ > 50)
```

Occurrence data was obtained from GBIF using the GBIF occurrence API.
This deals with synonyms etc. (using the GBIF Backbone Taxonomy as a standard), so e.g. retrieving data for "Bolboschoenus" includes all species and subspecies, including specimens described as *Bolboschoenus sp.*, *Bolboschoenus maritimus* and *Scirpus maritimus*.

```{r fetch-occ}
# TODO: Refactor
gbif_data <- function(scientificName, ...) {
  gbif_data <- rgbif::occ_data(scientificName = scientificName, ...)
  gbif_data$data
}

flora |>
  mutate(
    occ = future_map(taxon, gbif_data, 
                     hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
                     geometry = w_asia, limit = 100000)
  ) ->
  flora
```

```{r tidy-occ}
# TODO: further tidying. Ideas:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
# - https://cran.r-project.org/web/packages/CoordinateCleaner/vignettes/Cleaning_GBIF_data_with_CoordinateCleaner.html
# - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
flora |>
  mutate(
    occ = future_map(occ, gbif_drop_imprecise_coords, threshold = 5000),
    occ = future_map(occ, gbif_drop_fossils),
    occ = future_map(occ, gbif_drop_duplicate_coords),
    occ = future_map(occ, gbif_standardise_names),
    occ = future_map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    })
  ) ->
  flora
```

```{r fig-occ-map}
flora |>
  unnest(occ) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  ggplot() + 
  facet_wrap(vars(taxon)) +
  layer_spatial(ne_land, fill = "#ffffff") +
  geom_sf(shape = ".") +
  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
           ylim = w_asia[c("ymin", "ymax")],
           crs = utm37n,
           default_crs = latlong)
```

@@

```{r generate-background}
# TODO: Refactor to work with nested occ
# TODO: maybe defer this to just before the model is actually fit 
#   (and discard the background sample)
n_background <- 10000

occ |>
  mutate(present = TRUE) |>
  group_by(taxon) |>
  nest(.key = "occ_data") |>
  mutate(bg_data = map(1:n(), \(i) { 
    background_sample(
      region = w_asia_land, 
      N = n_background,
      coord_x = "longitude", coord_y = "latitude",
      present = FALSE
    )
  })) |>
  mutate(data = map2(occ_data, bg_data, bind_rows)) |>
  select(taxon, data) |>
  unnest(data) |>
  st_sf() ->
  occ
```

### Predictor data

@@

TODO: Soil, terrain (incl. TWI, slope?)

@@

```{r model}
rf_predictors <- function() {
  c(any_of("present"), starts_with("bio"))
}

# Drop any rows with NA values in any of the predictor variables
# TODO: Where do these actually come from? There's usually only a handful.
occ <- drop_na(occ, rf_predictors())

@@

rf_model <- rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger",
    probability = TRUE,
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1)
  )

# Fit to each taxon
# TODO: parallelise
occ |>
  nest(.by = taxon) |>
  mutate(
    data = map(data, as_tibble),
    fit = map(data, \(data) {
      fit(rf_model, factor(present) ~ ., data = select(data, rf_pred_vars()))
    })) ->
  rf_fits

# TODO: not this
rf_fits$taxon <- taxa$taxon
```

The output of the model is probalistic.
However, this should not be understood as an actual probability of occurrence [@CITE], but more akin to habitat suitability.
To simplify interpretation, we can convert these predictions into binary presence/absence maps, a process called "thresholding".
We select the threshold value for each model individually, using MaxSSS (as recommended by [@LuiEtAl2013])
This also makes it possible to analyses the predictions together as an assemblage.

```{r thresholding}
# TODO: thresholding
# - Pick a threshold
# - Reclassify pred raster
# - Extract predicted classification (for cur) for occ
# - ...or predict back onto the occurrence data? do we actually need the rasters for this?
# - Calculate sensitivity and with yardstick
# - Sum (SSS)
# - (Plot?)
# - Pick max SSS
```

@@

```{r predict}
# TODO: tidy
rf_fit <- filter(rf_fits, taxon == "Triticum urartu") |>
  pull(fit)
rf_fit <- rf_fit[[1]]

# TODO: Maybe combine stars and collapse period into dimension first?
rf_pred <- map(climate, predict, model = rf_fit, type = "prob")
rf_pred <- map(rf_pred, \(x) x[".pred_TRUE"])
rf_pred <- do.call(c, rf_pred)
rf_pred <- st_redimension(rf_pred, name = "period")

ggplot() + 
  geom_stars(data = st_crop(rf_pred, w_asia)) + 
  facet_wrap(~period) +
  scale_fill_batlow(reverse = TRUE)
```

### Model assessment & verification

```{r todo}
# TODO: model performance assessment:
# - R2
# - Boyce index
# - AUC, ROC, AUC-ROC, AUC-PRG ?
# - TSS

```
